Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-04-14 paper/intro.tex
second pass over intro/related
second pass related

There are prior studies of relaxed memory that include sequential
composition.  \citet{DBLP:journals/corr/abs-1804-04214} define a semantics
using pomsets without preconditions. Instead, their model uses syntactic
dependencies, thus invalidating many compiler optimizations.  They also
require a fence after every relaxed read on \armeight{}.

\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
calculating dependencies using event structures rather than logic.  They give
the semantics of sequential composition in continuation passing style,
whereas we give it in direct style.  The use step-indexing to account for
loops; we expect that the same approach could be applied here.

\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} also use event structures to
calculate dependencies, combined with an operational semantics that
incorporates program transformations.  This approach seems to require
whole-program analysis.

Other studies of relaxed memory can be categorized by their approach to
dependency calculation.  Hardware models use syntactic dependencies
\cite{alglave}.  Many software models do not bother with dependencies at all
\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
disallow compiler optimizations and efficient implementation, typically
requiring fences for every relaxed read on Arm
\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
  Dolan:2018:BDR:3192366.3192421, DBLP:conf/pldi/LahavVKHD17,
  DBLP:conf/lics/JeffreyR16, Boehm:2014:OGA:2618128.2618134,
  DBLP:journals/corr/abs-1804-04214}.

Many of the most prominent models are based on speculative execution
\cite{Manson:2005:JMM:1047659.1040336, DBLP:conf/esop/JagadeesanPR10,
  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
In their introduction, \citet{DBLP:journals/pacmpl/JagadeesanJR20} note that
these models fail to validate compositional reasoning of temporal
properties---see their examples OOTA4 and OOTA5 (from
\citep{DBLP:journals/toplas/Lochbihler13}).  The difference with our model
can be understood in terms of the valid program transformations.  The
speculative models allow reads to be introduced, with subsequent case
analysis on the value read---effectively, this can turn one read into two,
with different choices taken for the values read.  Our model invalidates this
transformation.  In return, our model enjoys compositionality for temporal
safety properties.

>>>>

There are prior studies of relaxed memory that include sequential
composition and/or precise calculation of semantic dependencies.
\citet{DBLP:conf/esop/PaviottiCPWOB20} give a denotational semantics,
calculating dependencies using event structures rather than logic.  They give
the semantics of sequential composition in continuation passing style,
whereas we give it in direct style.
They use step-indexing to account for
loops; we expect that the same approach could be applied here.
%
\citet{DBLP:journals/corr/abs-1804-04214} define a semantics
using pomsets without preconditions. Instead, their model uses syntactic
dependencies, thus invalidating many compiler optimizations.  They also
require a fence after every relaxed read on \armeight{}.
%
\citet{Pichon-Pharabod:2016:CSR:2837614.2837616} use event structures to
calculate dependencies, combined with an operational semantics that
incorporates program transformations.  This approach seems to require
whole-program analysis.

Other studies of relaxed memory can be categorized by their approach to
dependency calculation.  Hardware models use syntactic dependencies
\cite{alglave}.  Many software models do not bother with dependencies at all
\cite{Batty:2011:MCC:1926385.1926394, DBLP:journals/pacmpl/WattRP19,
  DBLP:conf/pldi/WattPPBDFPG20, goMM}.  Others have strong dependencies that
disallow compiler optimizations and efficient implementation, typically
requiring fences for every relaxed read on Arm
\cite{Lamport:1979:MMC:1311099.1311750, DBLP:conf/pldi/LahavVKHD17,
  Dolan:2018:BDR:3192366.3192421, DBLP:conf/lics/JeffreyR16,
  Boehm:2014:OGA:2618128.2618134}. %, DBLP:journals/corr/abs-1804-04214}.

Many of the most prominent models are based on speculative execution
\cite{Manson:2005:JMM:1047659.1040336, DBLP:conf/esop/JagadeesanPR10,
  DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}.
In their introduction, \citet{DBLP:journals/pacmpl/JagadeesanJR20} note that
these models fail to validate compositional reasoning of temporal
properties---see their examples OOTA4 and OOTA5 (from
\citep{DBLP:journals/toplas/Lochbihler13}).  The difference with our model
can be understood in terms of the valid program transformations.  The
speculative models allow reads to be introduced, with subsequent case
analysis on the value read---effectively, this can turn one read into two,
with different conditional branches taken for the two copies of the read.
Our model invalidates this transformation.  In return, our model enjoys
compositionality for temporal safety properties.
